---
output: github_document
---

# mgcvST

`mgcvST` provides proper Matérn SPDE smooths for `mgcv`, finite-element mesh
construction, and covariance score tests for pairs of spatial features. The
spatial range parameter `kappa` is fixed by design. An intercept is projected
from the SPDE coefficient space before fitting.

The primary analysis uses frequentist `mgcv` estimation, including `bam` for
larger data. INLA supplies a supplementary sparse SPDE estimator. Paired
validation holds the data, geometry and model terms fixed, then compares
scientific outputs while retaining estimator differences and convergence
diagnostics. See the [paired validation report](inst/notes/inla-bam-validation.md)
and [recorded results](inst/validation/inla-bam/).

The package provides two `mgcv` smooths:

- `bs = "spde"` fits the full projected SPDE basis.
- `bs = "spdePC"` eigendecomposes the fixed-`kappa` projected covariance
  `G = Q^{-1}` and retains the leading directions specified by
  `xt$pc_cutoff`. The fitted smooth stores `pc_score_Q`, the positive
  semidefinite score precision in the full projected coordinates, so that
  `score_test()` tests the original score space after truncation.

Two small spatial-transcriptomics data sets are included: `MISO_E13` and
`Visium_B`. Each contains covariates, three selected genes, a baseline mesh,
and a finer mesh for the `0.999` principal-component analysis. All figures
and results below are generated when this document is rendered.

## Installation

`mgcvST` uses CRAN, Bioconductor, and INLA dependencies. The high-throughput
`mgcvST.estimate()` workflow also uses the corrected marginal score from
`mgcv.taps`; the fitted-field figures below use `mgcViz`.

```{r installation, eval=FALSE}
install.packages("remotes")
install.packages("BiocManager")
BiocManager::install("BiocParallel", ask = FALSE)

install.packages(
  c("mgcViz", "INLA"),
  repos = c(
    getOption("repos"),
    INLA = "https://inla.r-inla-download.org/R/stable"
  )
)

remotes::install_github("harryyiheyang/mgcv.taps")
remotes::install_github("harryyiheyang/mgcvST", dependencies = TRUE)
```

## WGCNA modules from fitted score covariance

`mgcvST.wgcna()` identifies modules within gene blocks selected for downstream
analysis. It accepts a fitted `mgcvST.estimate()` or `inlaST.estimate()` result
and preserves the current fit's score definition, including its nuisance
adjustment. Each block uses the uncentered score covariance `crossprod(A) / q`,
where `q` is the number of selected score coordinates, followed by correlation,
WGCNA adjacency, topological overlap and dynamic tree cutting.

```r
install.packages(c("WGCNA", "dynamicTreeCut", "fastcluster"))

# fit is an existing estimate; genes contains the selected feature IDs.
W <- mgcvST.wgcna(fit, indices = genes)
W$modules
W$networks$selected$TOM

# Analyze previously selected blocks separately, preserving their gene order.
W <- mgcvST.wgcna(fit, indices = list(block1 = genes1, block2 = genes2),
                   group = "global", wgcna.para = list(power = 6))
```

The defaults are signed adjacency, power 6, signed TOM, average linkage,
minimum module size 20 and `deepSplit = 1`. A named `wgcna.para` list overrides
only the supplied settings. Specify `group` when a fit contains multiple
score components. Module labels are local to each gene block; label zero
indicates an unassigned gene. The result retains score coordinates, covariance,
correlation, adjacency, TOM and the clustering tree for further analysis.

See `help("mgcvST.wgcna")` for the interface and
[the source reconciliation record](inst/notes/wgcna-integration.md) for the
relationship between the earlier local WGCNA implementation and this release.

## INLA estimation with mandatory mean-zero spatial fields

`inlaST.set()` and `inlaST.estimate()` provide an INLA estimator with the
existing marginal and pairwise score-test interfaces. Every spatial field
is constrained to average zero over the model observations, including
each component of a global/local model.

```r
library(mgcvST)
data(MISO_E13, package = "mgcvST")
d <- MISO_E13$covariates
basis <- spde_basis(MISO_E13$meshes$spde, as.matrix(d[c("x", "y")]),
                    kappa = 0.1, project_intercept = TRUE)
model <- inlaST.set(
  response ~ offset(offset0) + s(x, y, bs = "spde", xt = basis),
  d, family = mgcv::nb(), precision_scale = "observation",
  control = list(control.inla = list(tolerance = 1e-4))
)
Y <- t(as.matrix(MISO_E13$expression))
fit <- inlaST.estimate(Y, model, retain_marginal = TRUE,
                       marginal_args = list(method = "liu"))
pairs <- t(combn(rownames(Y), 2L))
result <- inlaST.test(fit, pairs = pairs, calibration = "liu")
marginal <- inlaST.marginal(fit, calibration = "liu")
```

The backend uses fixed `kappa`, full SPDE bases and INLA empirical-Bayes
estimates. Spatial log precision and NB log size default to improper flat
objectives; Gaussian observation log precision defaults to
\(N(0,3^2)\). Each prior can be replaced explicitly with a validated INLA
prior. INLA and mgcv REML use distinct estimation procedures. Feature estimation uses
raw sparse precision matrices. The default score backend also uses sparse
precision solves for a single global SPDE; shared model setup and marginal
testing retain dense geometry. See [implementation and statistical details](inst/notes/inla-estimator.md),
the [runnable data example](inst/examples/inla_estimator.R), and the
[benchmark script](inst/benchmarks/inla-estimator.R).

The nuisance `Vp` now comes directly from INLA's constrained conditional
posterior covariance; the previous expected-Fisher reconstruction is retained
separately when `diagnostics = TRUE`; default fits skip that extra solve.
The workflow remains experimental. In independent
500-pair low-count NB validation (mean 0.3, size 2, 200 observations,
36 nodes, `kappa = 6`), nominal 5% rejection was 14.4% on the original
precision scale and 0.8% with `precision_scale = "observation"`, using native
posterior `Vp` and the conditioned SPDE score kernel. The latter was
substantially conservative, not exactly calibrated at 5%. These historical
validation figures used normal \(N(0,3^2)\) priors on
spatial log precision and NB log size, rather than the current flat defaults.

The example explicitly uses observation-scale precision: the spatial flat
log-prior is applied after normalizing the constrained field to unit average
observation variance. This changes the parameterization and optimizer start
on the FEM multiplier.
`precision_scale = "raw"` remains the default. Fixed precisions and returned
smoothing multipliers retain original FEM units. See the
[mechanism investigation](inst/notes/inla-lowcount-investigation.md) and
[earlier simulation interpretation](inst/notes/inla-estimated-null-interpretation.md).
`inlaST.estimate(..., score_backend = "auto")` now uses an algebraically
equivalent sparse score path for a single global SPDE with fixed nuisance
terms. Use `score_backend = "dense"` for comparisons. Models with additional
random blocks retain the dense path. The separate [raw-kernel research
helper](inst/benchmarks/inla-raw-kernel-sparse.R) is not the production method.

The complete formula above is prepared through the same mgcv geometry used by
`mgcvST.set()`. Existing code may instead supply the legacy `basis`
argument, or pass a prepared design as `inlaST.set(G = prepared$G)`.
Estimator controls supplied to `inlaST.set(..., control = ...)` are stored
in the model; `inlaST.estimate(..., control = ...)` overrides named values.
Prior specifications are replaced as complete objects. For example, the
historical normal prior can be requested with
`precision_prior = list(prior = "normal", param = c(0, 1/9))`.
For a global/local complete formula, the two SPDE terms cannot reuse exactly
the same coordinate term names because mgcv then creates duplicate
coefficient names. Use equal-valued aliases such as `u_local` and
`v_local` for the second term, or supply the legacy named basis list.

Flat priors are improper hyperpriors used for empirical-Bayes optimization;
they do not certify a proper hyperparameter posterior or an interior maximum.
Mean-zero constraints remain mandatory. See the
[flat-prior and large-spot investigation](inst/notes/inla-flat-prior.md).

The production score uses the SPDE kernel already conditioned on the
observation-mean constraint. Centering that conditioned kernel again does not
change it. The nuisance adjustment combines native INLA `Vp` with the
expected-likelihood working state, so its resulting operator need not satisfy
`P %*% 1 = 0` numerically. This does not make the conditioned kernel
equivalent to either the raw kernel or a separately formed
`C %*% G_raw %*% C`.

## Included data and observed spatial locations

```{r setup, message=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README_files/figure-gfm/"
)

library(mgcv)
library(mgcViz)
library(mgcvST)

data(MISO_E13, package = "mgcvST")
data(Visium_B, package = "mgcvST")

D_miso <- MISO_E13$covariates
D_visium <- Visium_B$covariates

data.frame(
  slice = c("MISO_E13", "Visium_B"),
  spots = c(nrow(D_miso), nrow(D_visium)),
  genes = c(
    paste(names(MISO_E13$expression), collapse = ", "),
    paste(names(Visium_B$expression), collapse = ", ")
  )
)
```

The package stores the original spatial coordinates and expression at each
included spot. The maps below display all observed spots, coloured by the raw
count of one selected gene in each slice.

```{r observed-spots, fig.width=10, fig.height=4.5, fig.cap="Observed spots in the two bundled slices. Colours show raw counts for the named gene.", fig.alt="Spatial spot maps for the MISO E13 and Visium B examples."}
col_miso <- hcl.colors(100L, "YlOrRd")[cut(
  log1p(MISO_E13$expression$Mapt), breaks = 100L, include.lowest = TRUE
)]
col_visium <- hcl.colors(100L, "YlOrRd")[cut(
  log1p(Visium_B$expression$mt_co3), breaks = 100L, include.lowest = TRUE
)]

op <- par(mfrow = c(1L, 2L), mar = c(4, 4, 3, 1))
plot(
  D_miso$x, D_miso$y, pch = 16L, cex = 0.55, col = col_miso,
  asp = 1, xlab = "x", ylab = "y", main = "MISO E13: Mapt"
)
plot(
  D_visium$x, D_visium$y, pch = 16L, cex = 0.55, col = col_visium,
  asp = 1, xlab = "x", ylab = "y", main = "Visium B: mt_co3"
)
par(op)
```

## MISO E13: PC-truncated SPDE fit, random field, and score test

The following chunk fits `Mapt` and `Map1b` on the finer MISO mesh. It fixes
`kappa = 0.1`, estimates only the smoothing parameter by REML, and retains
the `0.999` covariance-trace PC basis. `score_test()` uses the stored
positive-semidefinite `pc_score_Q` in the full projected coordinate system.

```{r miso-pc-fit-and-score}
D_miso_mapt <- D_miso
D_miso_map1b <- D_miso
D_miso_mapt$response <- MISO_E13$expression$Mapt
D_miso_map1b$response <- MISO_E13$expression$Map1b

fit_miso_mapt <- gam(
  response ~ offset(offset0) +
    s(x, y, bs = "spdePC", xt = list(
      mesh = MISO_E13$meshes$spdePC_g999,
      pc_cutoff = 0.999
    ), sp = c(-1, 0.1)),
  data = D_miso_mapt, family = nb(link = "log"), method = "REML"
)
fit_miso_map1b <- gam(
  response ~ offset(offset0) +
    s(x, y, bs = "spdePC", xt = list(
      mesh = MISO_E13$meshes$spdePC_g999,
      pc_cutoff = 0.999
    ), sp = c(-1, 0.1)),
  data = D_miso_map1b, family = nb(link = "log"), method = "REML"
)

sm_miso <- fit_miso_mapt$smooth[[1L]]
data.frame(
  full_projected_dimension = sm_miso$pc_full_dimension,
  retained_dimension = sm_miso$pc_retained_dimension,
  retained_trace = sm_miso$pc_cumulative_trace[sm_miso$pc_retained_dimension],
  score_Q_rank = as.integer(Matrix::rankMatrix(sm_miso$pc_score_Q))
)

score_miso <- score_test(fit_miso_mapt, fit_miso_map1b, method = "liu")
data.frame(
  pair = "Mapt--Map1b",
  signed_score = score_miso$signed_score,
  quadratic_statistic = score_miso$quadratic_statistic,
  information = score_miso$information,
  calibration = score_miso$method,
  p_value = score_miso$p_value
)
```

`mgcViz` evaluates the fitted random field on a regular grid through the
registered `spdePC` prediction matrix.

```{r miso-random-field, fig.width=7, fig.height=5, fig.cap="MISO E13 Mapt fitted PC-truncated SPDE random field from mgcViz.", fig.alt="Fitted spatial random field for Mapt in MISO E13."}
viz_miso <- getViz(fit_miso_mapt)
p_miso <- plot(sm(viz_miso, 1L)) +
  l_fitRaster() + l_fitContour() +
  ggplot2::labs(title = "MISO E13: Mapt fitted random field")
print(p_miso)
```

## Visium B: PC-truncated SPDE fit, random field, and score test

The Visium example uses the same fixed-`kappa` and `0.999` PC construction on
its finer mesh. Here we test `mt_co3` against `BRAFhuman`.

```{r visium-pc-fit-and-score}
D_visium_mtco3 <- D_visium
D_visium_braf <- D_visium
D_visium_mtco3$response <- Visium_B$expression$mt_co3
D_visium_braf$response <- Visium_B$expression$BRAFhuman

fit_visium_mtco3 <- gam(
  response ~ offset(offset0) +
    s(x, y, bs = "spdePC", xt = list(
      mesh = Visium_B$meshes$spdePC_g999,
      pc_cutoff = 0.999
    ), sp = c(-1, 0.1)),
  data = D_visium_mtco3, family = nb(link = "log"), method = "REML"
)
fit_visium_braf <- gam(
  response ~ offset(offset0) +
    s(x, y, bs = "spdePC", xt = list(
      mesh = Visium_B$meshes$spdePC_g999,
      pc_cutoff = 0.999
    ), sp = c(-1, 0.1)),
  data = D_visium_braf, family = nb(link = "log"), method = "REML"
)

sm_visium <- fit_visium_mtco3$smooth[[1L]]
data.frame(
  full_projected_dimension = sm_visium$pc_full_dimension,
  retained_dimension = sm_visium$pc_retained_dimension,
  retained_trace = sm_visium$pc_cumulative_trace[sm_visium$pc_retained_dimension],
  score_Q_rank = as.integer(Matrix::rankMatrix(sm_visium$pc_score_Q))
)

score_visium <- score_test(
  fit_visium_mtco3, fit_visium_braf, method = "liu"
)
data.frame(
  pair = "mt_co3--BRAFhuman",
  signed_score = score_visium$signed_score,
  quadratic_statistic = score_visium$quadratic_statistic,
  information = score_visium$information,
  calibration = score_visium$method,
  p_value = score_visium$p_value
)
```

```{r visium-random-field, fig.width=7, fig.height=5, fig.cap="Visium B mt_co3 fitted PC-truncated SPDE random field from mgcViz. Grey polygons are the two interior holes excluded by the manual SPDE boundary.", fig.alt="Fitted spatial random field for mt_co3 in Visium B, with the two excluded interior holes shown in grey."}
viz_visium <- getViz(fit_visium_mtco3)
visium_rings <- sf::st_geometry(
  Visium_B$meshes$spdePC_g999$domain.original
)[[1L]]
visium_holes <- do.call(rbind, lapply(
  seq_along(visium_rings[-1L]),
  function(i) data.frame(
    x = visium_rings[[i + 1L]][, 1L],
    y = visium_rings[[i + 1L]][, 2L],
    hole = i
  )
))
p_visium <- plot(sm(viz_visium, 1L)) +
  l_fitRaster() + l_fitContour() +
  ggplot2::geom_polygon(
    data = visium_holes,
    mapping = ggplot2::aes(x = x, y = y, group = hole),
    inherit.aes = FALSE, fill = "grey85", colour = "grey65",
    linewidth = 0.25, show.legend = FALSE
  ) +
  ggplot2::labs(title = "Visium B: mt_co3 fitted random field")
print(p_visium)
```

## High-throughput baseline-SPDE workflow

`mgcvST.estimate()` and `mgcvST.test()` retain only the compact summaries
needed for a pre-specified pair universe. Their current high-throughput
implementation uses the positive-definite baseline `spde` precision. The
following executable chunk fits all three MISO E13 genes and tests all three
pairs; no full `gam` objects are retained after estimation.

```{r miso-high-throughput}
genes_miso <- c("Mapt", "Map1b", "Hist1h2ao")
D_miso_setup <- D_miso
D_miso_setup$response <- MISO_E13$expression[[genes_miso[1L]]]
G_miso <- gam(
  response ~ offset(offset0) +
    s(x, y, bs = "spde", xt = list(mesh = MISO_E13$meshes$spde),
      sp = c(-1, 0.1)),
  data = D_miso_setup, family = nb(link = "log"), method = "REML",
  fit = FALSE, control = gam.control(nthreads = 1L, ncv.threads = 1L)
)
Y_miso <- t(as.matrix(MISO_E13$expression[, genes_miso, drop = FALSE]))

fit_batch_miso <- mgcvST.estimate(
  Y_miso, G_miso, feature_id = genes_miso,
  BPPARAM = BiocParallel::SerialParam()
)
test_batch_miso <- mgcvST.test(
  fit_batch_miso, pairs = t(combn(genes_miso, 2L)),
  BPPARAM = BiocParallel::SerialParam(), calibration = "liu", threads = 1L
)
test_batch_miso$results[, c(
  "feature1", "feature2", "score", "p_value", "p_adjusted", "status"
)]
```

For the packaged three-core versions, set an output directory and source the
two scripts below. Each script writes fit diagnostics, pair results, and
compact R objects to `MGCVST_EXAMPLE_OUTPUT`.

```{r packaged-data-examples, eval=FALSE}
dir.create("mgcvST-results", showWarnings = FALSE)
Sys.setenv(
  MGCVST_EXAMPLE_OUTPUT = normalizePath("mgcvST-results")
)

source(system.file("examples", "MISO_E13.R", package = "mgcvST"))
source(system.file("examples", "Visium_B.R", package = "mgcvST"))
```

## Reproduce the bundled simulations

`inst/examples/null.R` performs the null tail/type-I simulation. Its defaults
are 100,000 simulations and 20 workers. `inst/examples/alternative.R` uses
100 replicates at each of 100 correlation values. Both compare the baseline
and `0.999` PC-truncated representations on the included MISO E13 and Visium
B examples and save their results under `MGCVST_EXAMPLE_OUTPUT`.

For a quick local smoke run, reduce the counts before sourcing either script.

```{r simulations, eval=FALSE}
Sys.setenv(
  MGCVST_EXAMPLE_OUTPUT = normalizePath("mgcvST-results"),
  MGCVST_EXAMPLE_WORKERS = 3,
  MGCVST_EXAMPLE_NULL_REPS = 300,
  MGCVST_EXAMPLE_POWER_REPS = 10
)

source(system.file("examples", "null.R", package = "mgcvST"))
source(system.file("examples", "alternative.R", package = "mgcvST"))
```

## Construct a mesh from a boundary

`spde_mesh()` accepts either a manually curated polygon boundary or a boundary
returned by `spde_boundary_learn()`. The example data use curated boundaries.

```{r mesh, eval=FALSE}
loc <- as.matrix(MISO_E13$covariates[, c("x", "y")])
boundary <- spde_boundary_learn(loc, n_holes = 0L)
mesh <- spde_mesh(boundary, loc = loc, target = 150L)
plot(mesh)
```
