---
output: github_document
---

# mgcvST

`mgcvST` provides proper Matérn SPDE smooths for `mgcv`, finite-element mesh
construction, and covariance score tests for pairs of spatial features. The
spatial range parameter `kappa` is fixed by design. An intercept is projected
from the SPDE coefficient space before fitting.

The package provides two `mgcv` smooths:

- `bs = "spde"` fits the full projected SPDE basis.
- `bs = "spdePC"` eigendecomposes the fixed-`kappa` projected covariance
  `G = Q^{-1}` and retains the leading directions specified by
  `xt$pc_cutoff`. The fitted smooth stores `pc_score_Q`, the positive
  semidefinite score precision in the full projected coordinates, so that
  `score_test()` tests the original score space after truncation.

Two small spatial-transcriptomics data sets are included: `MISO_E13` and
`Visium_B`. Each contains covariates, three selected genes, a baseline mesh,
and a finer mesh for the `0.999` principal-component analysis. All figures
and results below are generated when this document is rendered.

## Installation

`mgcvST` uses CRAN, Bioconductor, and INLA dependencies. The high-throughput
`mgcvST.estimate()` workflow also uses the corrected marginal score from
`mgcv.taps`; the fitted-field figures below use `mgcViz`.

```{r installation, eval=FALSE}
install.packages("remotes")
install.packages("BiocManager")
BiocManager::install("BiocParallel", ask = FALSE)

install.packages(
  c("mgcViz", "INLA"),
  repos = c(
    getOption("repos"),
    INLA = "https://inla.r-inla-download.org/R/stable"
  )
)

remotes::install_github("harryyiheyang/mgcv.taps")
remotes::install_github("harryyiheyang/mgcvST", dependencies = TRUE)
```

## Included data and observed spatial locations

```{r setup, message=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README_files/figure-gfm/"
)

library(mgcv)
library(mgcViz)
library(mgcvST)

data(MISO_E13, package = "mgcvST")
data(Visium_B, package = "mgcvST")

D_miso <- MISO_E13$covariates
D_visium <- Visium_B$covariates

data.frame(
  slice = c("MISO_E13", "Visium_B"),
  spots = c(nrow(D_miso), nrow(D_visium)),
  genes = c(
    paste(names(MISO_E13$expression), collapse = ", "),
    paste(names(Visium_B$expression), collapse = ", ")
  )
)
```

The package stores the original spatial coordinates and expression at each
included spot. The maps below display all observed spots, coloured by the raw
count of one selected gene in each slice.

```{r observed-spots, fig.width=10, fig.height=4.5, fig.cap="Observed spots in the two bundled slices. Colours show raw counts for the named gene.", fig.alt="Spatial spot maps for the MISO E13 and Visium B examples."}
col_miso <- hcl.colors(100L, "YlOrRd")[cut(
  log1p(MISO_E13$expression$Mapt), breaks = 100L, include.lowest = TRUE
)]
col_visium <- hcl.colors(100L, "YlOrRd")[cut(
  log1p(Visium_B$expression$mt_co3), breaks = 100L, include.lowest = TRUE
)]

op <- par(mfrow = c(1L, 2L), mar = c(4, 4, 3, 1))
plot(
  D_miso$x, D_miso$y, pch = 16L, cex = 0.55, col = col_miso,
  asp = 1, xlab = "x", ylab = "y", main = "MISO E13: Mapt"
)
plot(
  D_visium$x, D_visium$y, pch = 16L, cex = 0.55, col = col_visium,
  asp = 1, xlab = "x", ylab = "y", main = "Visium B: mt_co3"
)
par(op)
```

## MISO E13: PC-truncated SPDE fit, random field, and score test

The following chunk fits `Mapt` and `Map1b` on the finer MISO mesh. It fixes
`kappa = 0.1`, estimates only the smoothing parameter by REML, and retains
the `0.999` covariance-trace PC basis. `score_test()` uses the stored
positive-semidefinite `pc_score_Q` in the full projected coordinate system.

```{r miso-pc-fit-and-score}
D_miso_mapt <- D_miso
D_miso_map1b <- D_miso
D_miso_mapt$response <- MISO_E13$expression$Mapt
D_miso_map1b$response <- MISO_E13$expression$Map1b

fit_miso_mapt <- gam(
  response ~ offset(offset0) +
    s(x, y, bs = "spdePC", xt = list(
      mesh = MISO_E13$meshes$spdePC_g999,
      pc_cutoff = 0.999
    ), sp = c(-1, 0.1)),
  data = D_miso_mapt, family = nb(link = "log"), method = "REML"
)
fit_miso_map1b <- gam(
  response ~ offset(offset0) +
    s(x, y, bs = "spdePC", xt = list(
      mesh = MISO_E13$meshes$spdePC_g999,
      pc_cutoff = 0.999
    ), sp = c(-1, 0.1)),
  data = D_miso_map1b, family = nb(link = "log"), method = "REML"
)

sm_miso <- fit_miso_mapt$smooth[[1L]]
data.frame(
  full_projected_dimension = sm_miso$pc_full_dimension,
  retained_dimension = sm_miso$pc_retained_dimension,
  retained_trace = sm_miso$pc_cumulative_trace[sm_miso$pc_retained_dimension],
  score_Q_rank = as.integer(Matrix::rankMatrix(sm_miso$pc_score_Q))
)

score_miso <- score_test(fit_miso_mapt, fit_miso_map1b, method = "liu")
data.frame(
  pair = "Mapt--Map1b",
  signed_score = score_miso$signed_score,
  quadratic_statistic = score_miso$quadratic_statistic,
  information = score_miso$information,
  calibration = score_miso$method,
  p_value = score_miso$p_value
)
```

`mgcViz` evaluates the fitted random field on a regular grid through the
registered `spdePC` prediction matrix.

```{r miso-random-field, fig.width=7, fig.height=5, fig.cap="MISO E13 Mapt fitted PC-truncated SPDE random field from mgcViz.", fig.alt="Fitted spatial random field for Mapt in MISO E13."}
viz_miso <- getViz(fit_miso_mapt)
p_miso <- plot(sm(viz_miso, 1L)) +
  l_fitRaster() + l_fitContour() +
  ggplot2::labs(title = "MISO E13: Mapt fitted random field")
print(p_miso)
```

## Visium B: PC-truncated SPDE fit, random field, and score test

The Visium example uses the same fixed-`kappa` and `0.999` PC construction on
its finer mesh. Here we test `mt_co3` against `BRAFhuman`.

```{r visium-pc-fit-and-score}
D_visium_mtco3 <- D_visium
D_visium_braf <- D_visium
D_visium_mtco3$response <- Visium_B$expression$mt_co3
D_visium_braf$response <- Visium_B$expression$BRAFhuman

fit_visium_mtco3 <- gam(
  response ~ offset(offset0) +
    s(x, y, bs = "spdePC", xt = list(
      mesh = Visium_B$meshes$spdePC_g999,
      pc_cutoff = 0.999
    ), sp = c(-1, 0.1)),
  data = D_visium_mtco3, family = nb(link = "log"), method = "REML"
)
fit_visium_braf <- gam(
  response ~ offset(offset0) +
    s(x, y, bs = "spdePC", xt = list(
      mesh = Visium_B$meshes$spdePC_g999,
      pc_cutoff = 0.999
    ), sp = c(-1, 0.1)),
  data = D_visium_braf, family = nb(link = "log"), method = "REML"
)

sm_visium <- fit_visium_mtco3$smooth[[1L]]
data.frame(
  full_projected_dimension = sm_visium$pc_full_dimension,
  retained_dimension = sm_visium$pc_retained_dimension,
  retained_trace = sm_visium$pc_cumulative_trace[sm_visium$pc_retained_dimension],
  score_Q_rank = as.integer(Matrix::rankMatrix(sm_visium$pc_score_Q))
)

score_visium <- score_test(
  fit_visium_mtco3, fit_visium_braf, method = "liu"
)
data.frame(
  pair = "mt_co3--BRAFhuman",
  signed_score = score_visium$signed_score,
  quadratic_statistic = score_visium$quadratic_statistic,
  information = score_visium$information,
  calibration = score_visium$method,
  p_value = score_visium$p_value
)
```

```{r visium-random-field, fig.width=7, fig.height=5, fig.cap="Visium B mt_co3 fitted PC-truncated SPDE random field from mgcViz. Grey polygons are the two interior holes excluded by the manual SPDE boundary.", fig.alt="Fitted spatial random field for mt_co3 in Visium B, with the two excluded interior holes shown in grey."}
viz_visium <- getViz(fit_visium_mtco3)
visium_rings <- sf::st_geometry(
  Visium_B$meshes$spdePC_g999$domain.original
)[[1L]]
visium_holes <- do.call(rbind, lapply(
  seq_along(visium_rings[-1L]),
  function(i) data.frame(
    x = visium_rings[[i + 1L]][, 1L],
    y = visium_rings[[i + 1L]][, 2L],
    hole = i
  )
))
p_visium <- plot(sm(viz_visium, 1L)) +
  l_fitRaster() + l_fitContour() +
  ggplot2::geom_polygon(
    data = visium_holes,
    mapping = ggplot2::aes(x = x, y = y, group = hole),
    inherit.aes = FALSE, fill = "grey85", colour = "grey65",
    linewidth = 0.25, show.legend = FALSE
  ) +
  ggplot2::labs(title = "Visium B: mt_co3 fitted random field")
print(p_visium)
```

## High-throughput baseline-SPDE workflow

`mgcvST.estimate()` and `mgcvST.test()` retain only the compact summaries
needed for a pre-specified pair universe. Their current high-throughput
implementation uses the positive-definite baseline `spde` precision. The
following executable chunk fits all three MISO E13 genes and tests all three
pairs; no full `gam` objects are retained after estimation.

```{r miso-high-throughput}
genes_miso <- c("Mapt", "Map1b", "Hist1h2ao")
D_miso_setup <- D_miso
D_miso_setup$response <- MISO_E13$expression[[genes_miso[1L]]]
G_miso <- gam(
  response ~ offset(offset0) +
    s(x, y, bs = "spde", xt = list(mesh = MISO_E13$meshes$spde),
      sp = c(-1, 0.1)),
  data = D_miso_setup, family = nb(link = "log"), method = "REML",
  fit = FALSE, control = gam.control(nthreads = 1L, ncv.threads = 1L)
)
Y_miso <- t(as.matrix(MISO_E13$expression[, genes_miso, drop = FALSE]))

fit_batch_miso <- mgcvST.estimate(
  Y_miso, G_miso, feature_id = genes_miso,
  BPPARAM = BiocParallel::SerialParam()
)
test_batch_miso <- mgcvST.test(
  fit_batch_miso, pairs = t(combn(genes_miso, 2L)),
  BPPARAM = BiocParallel::SerialParam(), calibration = "liu", threads = 1L
)
test_batch_miso$results[, c(
  "feature1", "feature2", "score", "p_value", "p_adjusted", "status"
)]
```

For the packaged three-core versions, set an output directory and source the
two scripts below. Each script writes fit diagnostics, pair results, and
compact R objects to `MGCVST_EXAMPLE_OUTPUT`.

```{r packaged-data-examples, eval=FALSE}
dir.create("mgcvST-results", showWarnings = FALSE)
Sys.setenv(
  MGCVST_EXAMPLE_OUTPUT = normalizePath("mgcvST-results")
)

source(system.file("examples", "MISO_E13.R", package = "mgcvST"))
source(system.file("examples", "Visium_B.R", package = "mgcvST"))
```

## Reproduce the bundled simulations

`inst/examples/null.R` performs the null tail/type-I simulation. Its defaults
are 100,000 simulations and 20 workers. `inst/examples/alternative.R` uses
100 replicates at each of 100 correlation values. Both compare the baseline
and `0.999` PC-truncated representations on the included MISO E13 and Visium
B examples and save their results under `MGCVST_EXAMPLE_OUTPUT`.

For a quick local smoke run, reduce the counts before sourcing either script.

```{r simulations, eval=FALSE}
Sys.setenv(
  MGCVST_EXAMPLE_OUTPUT = normalizePath("mgcvST-results"),
  MGCVST_EXAMPLE_WORKERS = 3,
  MGCVST_EXAMPLE_NULL_REPS = 300,
  MGCVST_EXAMPLE_POWER_REPS = 10
)

source(system.file("examples", "null.R", package = "mgcvST"))
source(system.file("examples", "alternative.R", package = "mgcvST"))
```

## Construct a mesh from a boundary

`spde_mesh()` accepts either a manually curated polygon boundary or a boundary
returned by `spde_boundary_learn()`. The example data use curated boundaries.

```{r mesh, eval=FALSE}
loc <- as.matrix(MISO_E13$covariates[, c("x", "y")])
boundary <- spde_boundary_learn(loc, n_holes = 0L)
mesh <- spde_mesh(boundary, loc = loc, target = 150L)
plot(mesh)
```
