---
output: github_document
---

# mgcvST

`mgcvST` provides proper Matérn SPDE smooths for `mgcv`, finite-element mesh
construction, and covariance score tests for pairs of spatial features. The
spatial range parameter `kappa` is fixed by design. An intercept is projected
from the SPDE coefficient space before fitting.

The primary analysis uses frequentist `mgcv` estimation, including `bam` for
larger data. INLA supplies a supplementary sparse SPDE estimator. Paired
validation holds the data, geometry and model terms fixed, then compares
scientific outputs while retaining estimator differences and convergence
diagnostics. See the [paired validation report](inst/notes/inla-bam-validation.md)
and [recorded results](inst/validation/inla-bam/).

The package provides two `mgcv` smooths:

- `bs = "spde"` fits the full projected SPDE basis.
- `bs = "spdePC"` eigendecomposes the fixed-`kappa` projected covariance
  `G = Q^{-1}` and retains the leading directions specified by
  `xt$pc_cutoff`. The fitted smooth stores `pc_score_Q`, the positive
  semidefinite score precision in the full projected coordinates, so that
  the raw-`G` workflow below tests the original score space after truncation.

Two small spatial-transcriptomics data sets are included: `MISO_E13` and
`Visium_B`. Each contains covariates, three selected genes, a baseline mesh,
and a finer mesh for the `0.999` principal-component analysis. Their figures
and results are regenerated when this document is rendered. The final MAGIC
figure is reproduced separately from the larger local data and saved fits.

## Installation

`mgcvST` uses CRAN, Bioconductor, and INLA dependencies. The high-throughput
`mgcvST.estimate()` workflow includes its corrected marginal score calculation;
the fitted-field figures below use `ggplot2`.

```{r installation, eval=FALSE}
install.packages("remotes")
install.packages("BiocManager")
BiocManager::install("BiocParallel", ask = FALSE)

install.packages(
  c("ggplot2", "INLA"),
  repos = c(
    getOption("repos"),
    INLA = "https://inla.r-inla-download.org/R/stable"
  )
)

remotes::install_github("harryyiheyang/mgcvST", dependencies = TRUE)
```

## WGCNA modules from fitted score covariance

`mgcvST.wgcna()` identifies modules within gene blocks selected for downstream
analysis. It accepts a fitted `mgcvST.estimate()` or `inlaST.estimate()` result
and preserves the current fit's score definition, including its nuisance
adjustment. Each block uses the uncentered score covariance `crossprod(A) / q`,
where `q` is the number of selected score coordinates, followed by correlation,
WGCNA adjacency, topological overlap and dynamic tree cutting.

```r
install.packages(c("WGCNA", "dynamicTreeCut", "fastcluster"))

# fit is an existing estimate; genes contains the selected feature IDs.
W <- mgcvST.wgcna(fit, indices = genes)
W$modules
W$networks$selected$TOM

# Analyze previously selected blocks separately, preserving their gene order.
W <- mgcvST.wgcna(fit, indices = list(block1 = genes1, block2 = genes2), wgcna.para = list(power = 6))
```

The defaults are signed adjacency, power 6, signed TOM, average linkage,
minimum module size 20 and `deepSplit = 1`. A named `wgcna.para` list overrides
only the supplied settings. For a fit from `inlaST.estimate()` use
`inlaST.wgcna()`, the sparse-kernel sibling with the same arguments and the
same downstream splitting; `mgcvST.wgcna()` also accepts an INLA fit and
dispatches to that same sparse kernel. Module labels are local to each gene
block; label zero indicates an unassigned gene. The result retains score coordinates, covariance,
correlation, adjacency, TOM and the clustering tree for further analysis.

See `help("mgcvST.wgcna")` for the interface and
[the source reconciliation record](inst/notes/wgcna-integration.md) for the
relationship between the earlier local WGCNA implementation and this release.

## INLA estimation with mandatory mean-zero spatial fields

`inlaST.set()` and `inlaST.estimate()` provide an INLA estimator with the
existing marginal and pairwise score-test interfaces. Every spatial field
is constrained to average zero over the model observations for the global
spatial score process.

```r
library(mgcvST)
data(MISO_E13, package = "mgcvST")
d <- MISO_E13$covariates
basis <- spde_basis(MISO_E13$meshes$spde, as.matrix(d[c("x", "y")]),
                    kappa = 0.1, project_intercept = TRUE)
model <- inlaST.set(
  response ~ offset(offset0) + s(x, y, bs = "spde", xt = basis),
  d, family = mgcv::nb(), precision_scale = "observation",
  control = list(control.inla = list(tolerance = 1e-4))
)
Y <- t(as.matrix(MISO_E13$expression))
fit <- inlaST.estimate(Y, model, retain_marginal = TRUE,
                       marginal_args = list(method = "liu"))
pairs <- t(combn(rownames(Y), 2L))
result <- inlaST.test(fit, pairs = pairs, calibration = "liu")
marginal <- inlaST.marginal(fit, calibration = "liu")
```

The backend uses fixed `kappa`, full SPDE bases and INLA empirical-Bayes
estimates. Spatial log precision, NB log size and Gaussian observation log
precision default to flat hyperparameter objectives. For Gaussian observations,
this is a prior on inverse residual variance; the observation distribution
remains Gaussian. Each prior can be replaced explicitly with a validated INLA
prior. INLA and mgcv REML use distinct estimation procedures. Feature estimation uses
raw sparse precision matrices. The default score backend also uses sparse
precision solves for a single global SPDE; shared model setup and marginal
testing retain dense geometry. See [implementation and statistical details](inst/notes/inla-estimator.md),
the [runnable data example](inst/examples/inla_estimator.R), and the
[benchmark script](inst/benchmarks/inla-estimator.R).

The nuisance `Vp` now comes directly from INLA's constrained conditional
posterior covariance; the previous expected-Fisher reconstruction is retained
separately when `diagnostics = TRUE`; default fits skip that extra solve.
The workflow remains experimental. In independent
500-pair low-count NB validation (mean 0.3, size 2, 200 observations,
36 nodes, `kappa = 6`), nominal 5% rejection was 14.4% on the original
precision scale and 0.8% with `precision_scale = "observation"`, using native
posterior `Vp` and the conditioned SPDE score kernel. The latter was
substantially conservative, not exactly calibrated at 5%. These historical
validation figures used normal \(N(0,3^2)\) priors on
spatial log precision and NB log size, rather than the current flat defaults.

The example explicitly uses observation-scale precision: the spatial flat
log-prior is applied after normalizing the constrained field to unit average
observation variance. This changes the parameterization and optimizer start
on the FEM multiplier.
`precision_scale = "raw"` remains the default. Fixed precisions and returned
smoothing multipliers retain original FEM units. See the
[mechanism investigation](inst/notes/inla-lowcount-investigation.md) and
[earlier simulation interpretation](inst/notes/inla-estimated-null-interpretation.md).
The INLA path uses the native sparse score kernel for a single global SPDE with
fixed nuisance terms. A model with additional random blocks is rejected by
`inlaST.set()` with the reason from the sparse capability gate. The separate
[raw-kernel research helper](inst/benchmarks/inla-raw-kernel-sparse.R) is not
the production method.

The complete formula above is prepared through the same mgcv geometry used by
`mgcvST.set()`. Existing code may instead supply the legacy `basis`
argument, or pass a prepared design as `inlaST.set(G = prepared$G)`.
Estimator controls supplied to `inlaST.set(..., control = ...)` are stored
in the model; `inlaST.estimate(..., control = ...)` overrides named values.
Prior specifications are replaced as complete objects. For example, the
historical normal prior can be requested with
`precision_prior = list(prior = "normal", param = c(0, 1/9))`.
The public model interface uses one global SPDE score process. Additional
smooth terms may be included as nuisance structure, but they are not additional
spatial score components.

Flat priors are improper hyperpriors used for empirical-Bayes optimization;
they do not certify a proper hyperparameter posterior or an interior maximum.
Mean-zero constraints remain mandatory. See the
[flat-prior and large-spot investigation](inst/notes/inla-flat-prior.md).

The production score uses the SPDE kernel already conditioned on the
observation-mean constraint. Centering that conditioned kernel again does not
change it. The nuisance adjustment combines native INLA `Vp` with the
expected-likelihood working state, so its resulting operator need not satisfy
`P %*% 1 = 0` numerically. This does not make the conditioned kernel
equivalent to either the raw kernel or a separately formed
`C %*% G_raw %*% C`.

## Included data and observed spatial locations

```{r setup, message=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README_files/figure-gfm/"
)

library(mgcv)
library(ggplot2)
library(mgcvST)

data(MISO_E13, package = "mgcvST")
data(Visium_B, package = "mgcvST")

D_miso <- MISO_E13$covariates
D_visium <- Visium_B$covariates

data.frame(
  slice = c("MISO_E13", "Visium_B"),
  spots = c(nrow(D_miso), nrow(D_visium)),
  genes = c(
    paste(names(MISO_E13$expression), collapse = ", "),
    paste(names(Visium_B$expression), collapse = ", ")
  )
)
```

The package stores the original spatial coordinates and expression at each
included spot. The maps below display all observed spots, coloured by the raw
count of one selected gene in each slice.

```{r observed-spots, fig.width=10, fig.height=4.8, fig.cap="All observed spots in the two bundled slices. Each horizontal scale shows linear raw counts for its named gene, with white at zero.", fig.alt="Spatial spot maps for the MISO E13 and Visium B examples, with horizontal legends below."}
p_obs_miso <- ggplot2::ggplot(D_miso, ggplot2::aes(x, y)) +
  ggplot2::geom_point(ggplot2::aes(colour = MISO_E13$expression$Mapt), size = 1.3) +
  ggplot2::coord_equal() +
  ggplot2::scale_colour_gradientn(
    colours = c("#FFFFFF", "#6BAED6", "#08306B"),
    limits = c(0, max(MISO_E13$expression$Mapt)), name = "Raw count",
    guide = ggplot2::guide_colourbar(direction = "horizontal", title.position = "left",
      barwidth = grid::unit(40, "mm"), barheight = grid::unit(2.5, "mm"))
  ) +
  ggplot2::theme_classic() +
  ggplot2::theme(legend.position = "bottom", legend.direction = "horizontal",
    legend.box = "horizontal", plot.margin = ggplot2::margin(6, 16, 6, 6)) +
  ggplot2::labs(title = "MISO E13: Mapt")
p_obs_visium <- ggplot2::ggplot(D_visium, ggplot2::aes(x, y)) +
  ggplot2::geom_point(ggplot2::aes(colour = Visium_B$expression$mt_co3), size = 0.9) +
  ggplot2::coord_equal() +
  ggplot2::scale_colour_gradientn(
    colours = c("#FFFFFF", "#6BAED6", "#08306B"),
    limits = c(0, max(Visium_B$expression$mt_co3)), name = "Raw count",
    guide = ggplot2::guide_colourbar(direction = "horizontal", title.position = "left",
      barwidth = grid::unit(40, "mm"), barheight = grid::unit(2.5, "mm"))
  ) +
  ggplot2::theme_classic() +
  ggplot2::theme(legend.position = "bottom", legend.direction = "horizontal",
    legend.box = "horizontal", plot.margin = ggplot2::margin(6, 16, 6, 6)) +
  ggplot2::labs(title = "Visium B: mt_co3")
grid::grid.newpage()
grid::pushViewport(grid::viewport(layout = grid::grid.layout(1, 2)))
print(p_obs_miso, vp = grid::viewport(layout.pos.row = 1, layout.pos.col = 1))
print(p_obs_visium, vp = grid::viewport(layout.pos.row = 1, layout.pos.col = 2))
grid::popViewport()
```

## MISO E13: PC-truncated SPDE fit, random field, and score test

The following chunk fits `Mapt` and `Map1b` on the finer MISO mesh. It fixes
`kappa = 0.1`, estimates only the smoothing parameter by REML, and retains
the `0.999` covariance-trace PC basis. For the pairwise test, the raw
`gam(..., fit = FALSE)` setup is passed directly to `mgcvST.estimate()`.
This path uses the stored positive-semidefinite `pc_score_Q` in the full
projected coordinate system.

```{r miso-pc-fit-and-score}
D_miso_mapt <- D_miso
D_miso_map1b <- D_miso
D_miso_mapt$response <- MISO_E13$expression$Mapt
D_miso_map1b$response <- MISO_E13$expression$Map1b
basis_miso_pc <- spde_basis(
  MISO_E13$meshes$spdePC_g999, as.matrix(D_miso[, c("x", "y")]),
  kappa = 0.1, pc_cutoff = 0.999
)

fit_miso_mapt <- gam(
  response ~ offset(offset0) +
    s(x, y, bs = "spdePC", xt = basis_miso_pc, sp = -1),
  data = D_miso_mapt, family = nb(link = "log"), method = "REML"
)
fit_miso_map1b <- gam(
  response ~ offset(offset0) +
    s(x, y, bs = "spdePC", xt = basis_miso_pc, sp = -1),
  data = D_miso_map1b, family = nb(link = "log"), method = "REML"
)

sm_miso <- fit_miso_mapt$smooth[[1L]]
data.frame(
  full_projected_dimension = sm_miso$pc_full_dimension,
  retained_dimension = sm_miso$pc_retained_dimension,
  retained_trace = sm_miso$pc_cumulative_trace[sm_miso$pc_retained_dimension],
  score_Q_rank = as.integer(Matrix::rankMatrix(sm_miso$pc_score_Q))
)

G_miso_pc <- gam(
  formula(fit_miso_mapt), data = D_miso_mapt,
  family = nb(link = "log"), method = "REML", fit = FALSE,
  control = gam.control(nthreads = 1L, ncv.threads = 1L)
)
Y_miso_pc <- t(as.matrix(MISO_E13$expression[, c("Mapt", "Map1b")]))
fit_pair_miso <- mgcvST.estimate(
  Y_miso_pc, G_miso_pc, BPPARAM = BiocParallel::SerialParam()
)
score_miso <- mgcvST.test(
  fit_pair_miso, pairs = matrix(c("Mapt", "Map1b"), nrow = 1L),
  calibration = "liu", FDR = FALSE,
  BPPARAM = BiocParallel::SerialParam(), threads = 1L
)
data.frame(
  pair = "Mapt--Map1b",
  signed_score = score_miso$results$signed_score,
  quadratic_statistic = score_miso$results$statistic,
  information = score_miso$results$information,
  calibration = score_miso$calibration,
  p_value = score_miso$results$p_two_sided
)
```

The fitted random field is evaluated on a regular grid through the registered
`spdePC` prediction matrix. Predictions are restricted to the intersection of
the supplied domain and its triangulation, preserving boundaries and holes.

```{r miso-prediction-grid, include=FALSE}
grid_miso <- expand.grid(
  x = seq(min(D_miso$x), max(D_miso$x), length.out = 100L),
  y = seq(min(D_miso$y), max(D_miso$y), length.out = 100L)
)
grid_miso$offset0 <- 0
mesh_miso <- MISO_E13$meshes$spdePC_g999
loc_miso <- sweep(as.matrix(grid_miso[, c("x", "y")]),
                  2L, mesh_miso$transform$center, "-") / mesh_miso$transform$scale
inside_miso <- !is.na(geometry::tsearchn(
  mesh_miso$mesh$loc[, 1:2], mesh_miso$mesh$graph$tv, loc_miso
)$idx) & lengths(sf::st_within(
  sf::st_as_sf(grid_miso, coords = c("x", "y"),
               crs = sf::st_crs(mesh_miso$domain.original)),
  mesh_miso$domain.original
)) > 0L
grid_miso$field <- NA_real_
grid_miso$field[inside_miso] <- as.numeric(predict(
  fit_miso_mapt, newdata = grid_miso[inside_miso, ],
  type = "terms", terms = fit_miso_mapt$smooth[[1L]]$label
))
```

```{r miso-random-field, fig.width=7, fig.height=5, fig.cap="MISO E13 Mapt fitted PC-truncated SPDE random field, evaluated inside the supplied mesh.", fig.alt="Fitted spatial random field for Mapt in MISO E13, with a horizontal legend below."}
p_miso <- ggplot2::ggplot(grid_miso, ggplot2::aes(x, y)) +
  ggplot2::geom_raster(ggplot2::aes(fill = field), na.rm = TRUE) +
  ggplot2::geom_contour(data = grid_miso, ggplot2::aes(x, y, z = field),
    inherit.aes = FALSE, colour = "black", linewidth = 0.2, na.rm = TRUE) +
  ggplot2::scale_fill_viridis_c(name = "Spatial field", na.value = "grey85") +
  ggplot2::coord_equal(expand = FALSE) +
  ggplot2::theme_classic() +
  ggplot2::labs(title = "MISO E13: Mapt fitted random field") +
  ggplot2::theme(legend.position = "bottom", legend.direction = "horizontal",
    legend.box = "horizontal") +
  ggplot2::guides(fill = ggplot2::guide_colourbar(direction = "horizontal",
    title.position = "left", barwidth = grid::unit(70, "mm"),
    barheight = grid::unit(2.5, "mm")),
    colour = ggplot2::guide_legend(nrow = 1))
print(p_miso)
```

## Visium B: PC-truncated SPDE fit, random field, and score test

The Visium example uses the same fixed-`kappa` and `0.999` PC construction on
its finer mesh. Here we test `mt_co3` against `BRAFhuman`.

```{r visium-pc-fit-and-score}
D_visium_mtco3 <- D_visium
D_visium_braf <- D_visium
D_visium_mtco3$response <- Visium_B$expression$mt_co3
D_visium_braf$response <- Visium_B$expression$BRAFhuman
basis_visium_pc <- spde_basis(
  Visium_B$meshes$spdePC_g999, as.matrix(D_visium[, c("x", "y")]),
  kappa = 0.1, pc_cutoff = 0.999
)

fit_visium_mtco3 <- gam(
  response ~ offset(offset0) +
    s(x, y, bs = "spdePC", xt = basis_visium_pc, sp = -1),
  data = D_visium_mtco3, family = nb(link = "log"), method = "REML"
)
fit_visium_braf <- gam(
  response ~ offset(offset0) +
    s(x, y, bs = "spdePC", xt = basis_visium_pc, sp = -1),
  data = D_visium_braf, family = nb(link = "log"), method = "REML"
)

sm_visium <- fit_visium_mtco3$smooth[[1L]]
data.frame(
  full_projected_dimension = sm_visium$pc_full_dimension,
  retained_dimension = sm_visium$pc_retained_dimension,
  retained_trace = sm_visium$pc_cumulative_trace[sm_visium$pc_retained_dimension],
  score_Q_rank = as.integer(Matrix::rankMatrix(sm_visium$pc_score_Q))
)

G_visium_pc <- gam(
  formula(fit_visium_mtco3), data = D_visium_mtco3,
  family = nb(link = "log"), method = "REML", fit = FALSE,
  control = gam.control(nthreads = 1L, ncv.threads = 1L)
)
Y_visium_pc <- t(as.matrix(Visium_B$expression[, c("mt_co3", "BRAFhuman")]))
fit_pair_visium <- mgcvST.estimate(
  Y_visium_pc, G_visium_pc, BPPARAM = BiocParallel::SerialParam()
)
score_visium <- mgcvST.test(
  fit_pair_visium, pairs = matrix(c("mt_co3", "BRAFhuman"), nrow = 1L),
  calibration = "liu", FDR = FALSE,
  BPPARAM = BiocParallel::SerialParam(), threads = 1L
)
data.frame(
  pair = "mt_co3--BRAFhuman",
  signed_score = score_visium$results$signed_score,
  quadratic_statistic = score_visium$results$statistic,
  information = score_visium$results$information,
  calibration = score_visium$calibration,
  p_value = score_visium$results$p_two_sided
)
```

```{r visium-prediction-grid, include=FALSE}
grid_visium <- expand.grid(
  x = seq(min(D_visium$x), max(D_visium$x), length.out = 100L),
  y = seq(min(D_visium$y), max(D_visium$y), length.out = 100L)
)
grid_visium$offset0 <- 0
mesh_visium <- Visium_B$meshes$spdePC_g999
loc_visium <- sweep(as.matrix(grid_visium[, c("x", "y")]),
                    2L, mesh_visium$transform$center, "-") / mesh_visium$transform$scale
inside_visium <- !is.na(geometry::tsearchn(
  mesh_visium$mesh$loc[, 1:2], mesh_visium$mesh$graph$tv, loc_visium
)$idx) & lengths(sf::st_within(
  sf::st_as_sf(grid_visium, coords = c("x", "y"),
               crs = sf::st_crs(mesh_visium$domain.original)),
  mesh_visium$domain.original
)) > 0L
grid_visium$field <- NA_real_
grid_visium$field[inside_visium] <- as.numeric(predict(
  fit_visium_mtco3, newdata = grid_visium[inside_visium, ],
  type = "terms", terms = fit_visium_mtco3$smooth[[1L]]$label
))
```

```{r visium-random-field, fig.width=7, fig.height=5, fig.cap="Visium B mt_co3 fitted PC-truncated SPDE random field. Grey polygons are the two interior holes excluded by the manual SPDE boundary.", fig.alt="Fitted spatial random field for mt_co3 in Visium B, with two excluded interior holes in grey and a horizontal legend below."}
visium_rings <- sf::st_geometry(
  Visium_B$meshes$spdePC_g999$domain.original
)[[1L]]
visium_holes <- do.call(rbind, lapply(
  seq_along(visium_rings[-1L]),
  function(i) data.frame(
    x = visium_rings[[i + 1L]][, 1L],
    y = visium_rings[[i + 1L]][, 2L],
    hole = i
  )
))
p_visium <- ggplot2::ggplot(grid_visium, ggplot2::aes(x, y)) +
  ggplot2::geom_raster(ggplot2::aes(fill = field), na.rm = TRUE) +
  ggplot2::geom_contour(data = grid_visium, ggplot2::aes(x, y, z = field),
    inherit.aes = FALSE, colour = "black", linewidth = 0.2, na.rm = TRUE) +
  ggplot2::scale_fill_viridis_c(name = "Spatial field", na.value = "grey85") +
  ggplot2::coord_equal(expand = FALSE) +
  ggplot2::theme_classic() +
  ggplot2::geom_polygon(
    data = visium_holes,
    mapping = ggplot2::aes(x = x, y = y, group = hole),
    inherit.aes = FALSE, fill = "grey85", colour = "grey65",
    linewidth = 0.25, show.legend = FALSE
  ) +
  ggplot2::labs(title = "Visium B: mt_co3 fitted random field") +
  ggplot2::theme(legend.position = "bottom", legend.direction = "horizontal",
    legend.box = "horizontal") +
  ggplot2::guides(fill = ggplot2::guide_colourbar(direction = "horizontal",
    title.position = "left", barwidth = grid::unit(70, "mm"),
    barheight = grid::unit(2.5, "mm")),
    colour = ggplot2::guide_legend(nrow = 1))
print(p_visium)
```

## High-throughput baseline-SPDE workflow

`mgcvST.estimate()` and `mgcvST.test()` retain only the compact summaries
needed for a pre-specified pair universe. Their current high-throughput
implementation uses the positive-definite baseline `spde` precision. The
following executable chunk fits all three MISO E13 genes and tests all three
pairs; no full `gam` objects are retained after estimation.

```{r miso-high-throughput}
genes_miso <- c("Mapt", "Map1b", "Hist1h2ao")
D_miso_setup <- D_miso
D_miso_setup$response <- MISO_E13$expression[[genes_miso[1L]]]
basis_miso_full <- spde_basis(
  MISO_E13$meshes$spde, as.matrix(D_miso_setup[, c("x", "y")]), kappa = 0.1
)
G_miso <- gam(
  response ~ offset(offset0) +
    s(x, y, bs = "spde", xt = basis_miso_full, sp = -1),
  data = D_miso_setup, family = nb(link = "log"), method = "REML",
  fit = FALSE, control = gam.control(nthreads = 1L, ncv.threads = 1L)
)
Y_miso <- t(as.matrix(MISO_E13$expression[, genes_miso, drop = FALSE]))

fit_batch_miso <- mgcvST.estimate(
  Y_miso, G_miso, feature_id = genes_miso,
  BPPARAM = BiocParallel::SerialParam()
)
test_batch_miso <- mgcvST.test(
  fit_batch_miso, pairs = t(combn(genes_miso, 2L)),
  BPPARAM = BiocParallel::SerialParam(), calibration = "liu", threads = 1L
)
test_batch_miso$results[, c(
  "feature1", "feature2", "signed_score", "p_two_sided", "p_adjusted"
)]
```

For the packaged three-core versions, set an output directory and source the
two scripts below. Each script writes fit diagnostics, pair results, and
compact R objects to `MGCVST_EXAMPLE_OUTPUT`.

```{r packaged-data-examples, eval=FALSE}
dir.create("mgcvST-results", showWarnings = FALSE)
Sys.setenv(
  MGCVST_EXAMPLE_OUTPUT = normalizePath("mgcvST-results")
)

source(system.file("examples", "MISO_E13.R", package = "mgcvST"))
source(system.file("examples", "Visium_B.R", package = "mgcvST"))
```

## Simulation reference scripts

`inst/examples/null.R` and `inst/examples/alternative.R` retain the original
null-tail and power simulation designs for the baseline and `0.999`
PC-truncated representations. These historical scripts use earlier smooth
and score interfaces and require migration before running with the current
release. See the [paired validation report](inst/notes/inla-bam-validation.md)
for the settings and results of the newer INLA–mgcv comparisons.

## Construct a mesh from a boundary

`spde_mesh()` accepts either a manually curated polygon boundary or a boundary
returned by `spde_boundary_learn()`. The example data use curated boundaries.

```{r mesh, eval=FALSE}
loc <- as.matrix(MISO_E13$covariates[, c("x", "y")])
boundary <- spde_boundary_learn(loc, n_holes = 0L)
mesh <- spde_mesh(boundary, loc = loc, target = 150L)
plot(mesh)
```

## Three-dimensional MAGIC example

The native INLA interface also supports tetrahedral meshes for stacked
sections. The MAGIC example contains 97,830 measured spots on 93 sections,
with their recorded, unequal z spacing retained. Its supplied mesh has
1,962 nodes and 7,676 tetrahedra. Tfap2b is one of the regional markers
displayed in Figure 5d of the [source paper](https://doi.org/10.1038/s41588-024-01906-4).

![MAGIC section stack, tetrahedral mesh and three-dimensional fitted Tfap2b
expression, shown as static projections with horizontal legends below.](man/figures/magic-3d-overview.png)

The panels show the observed section stack, the supplied mesh with its
nodes and triangular boundary faces, and fitted Tfap2b expression at the
observed locations. All panels use the same projection and physical scale.
The mesh view shows a cutaway of its original boundary faces (face-centre
x ≤ 4.72 mm), together with all nodes. Expression is displayed as a linear
rate per 10,000 UMI, starting at white for zero and increasing through blue;
values above the 98th percentile (4.62) share the darkest colour.

The fitted expression comes from the saved full-data, spatial-only INLA model
with flat hyperparameter priors and the historical integral constraint.
Section-level random effects remain an ongoing investigation. See the
[marker and section-effect exploration](inst/notes/magic-slide-exploration.md)
and [data construction instructions](inst/notes/magic-data.md).
Reproduce this figure with `inst/examples/magic-readme-3d.R` after preparing
the local MAGIC data and saved fits. A separate current-native-API simulation
is available in `inst/examples/inla3d-visualization-native-demo.R`.

The [3D feasibility study](inst/notes/inla3d-adaptive.md) records the
geometry-adaptive mesh experiments, and the
[transfer notes](inst/notes/inla3d-transfer.md) describe the real-data geometry.
The [flat-prior stress study](inst/notes/inla-stress-validation.md) and
[spectrum diagnostics](inst/notes/inla-score-spectrum.md) retain the
experimental score results and their subsequent sensitivity analyses.
