---
output: github_document
---

# mgcvST

`mgcvST` provides proper Matérn SPDE smooths for `mgcv`, finite-element mesh
construction, and covariance score tests for pairs of spatial features. The
spatial range parameter `kappa` is fixed by design. An intercept is projected
from the SPDE coefficient space before fitting.

The package provides two `mgcv` smooths:

- `bs = "spde"` fits the full projected SPDE basis.
- `bs = "spdePC"` eigendecomposes the fixed-`kappa` projected covariance
  `G = Q^{-1}` and retains the leading directions specified by
  `xt$pc_cutoff`. The fitted smooth stores `pc_score_Q`, the positive
  semidefinite score precision in the full projected coordinates, so that
  `score_test()` can test the original score space after truncation.

Two small spatial-transcriptomics data sets are included: `MISO_E13` and
`Visium_B`. Each includes covariates, three selected genes, a baseline mesh,
and a finer mesh for the `0.999` principal-component analysis.

## Installation

`mgcvST` uses CRAN, Bioconductor, and INLA dependencies. The high-throughput
`mgcvST.estimate()` workflow also uses the corrected marginal score from
`mgcv.taps`.

```{r installation, eval=FALSE}
install.packages("remotes")
install.packages("BiocManager")
BiocManager::install("BiocParallel", ask = FALSE)

install.packages(
  "INLA",
  repos = c(
    getOption("repos"),
    INLA = "https://inla.r-inla-download.org/R/stable"
  )
)

remotes::install_github("harryyiheyang/mgcv.taps")
remotes::install_github("harryyiheyang/mgcvST", dependencies = TRUE)
```

## Included data

```{r data, eval=FALSE}
library(mgcvST)

data(MISO_E13)
names(MISO_E13)
names(MISO_E13$expression)
# [1] "Mapt" "Map1b" "Hist1h2ao"

data(Visium_B)
names(Visium_B$expression)
# [1] "mt_co3" "mt_co2" "BRAFhuman"
```

For both objects, `covariates` contains `x`, `y`, `offset0`, and the original
spot identifier. The rows of `expression` and `covariates` are aligned.
`meshes$spde` is the baseline mesh. `meshes$spdePC_g999` is the finer mesh
used with `bs = "spdePC"` and `pc_cutoff = 0.999`.

## Fit two features and run a PC score test

This is the direct workflow for a selected pair. `sp = c(-1, 0.1)` estimates
only the smoothing parameter by REML and keeps `kappa = 0.1` fixed.

```{r direct-pc-pair, eval=FALSE}
library(mgcv)
library(mgcvST)

data(MISO_E13)
D1 <- MISO_E13$covariates
D2 <- MISO_E13$covariates
D1$response <- MISO_E13$expression$Mapt
D2$response <- MISO_E13$expression$Map1b

fit1 <- gam(
  response ~ offset(offset0) +
    s(x, y, bs = "spdePC", xt = list(
      mesh = MISO_E13$meshes$spdePC_g999,
      pc_cutoff = 0.999
    ), sp = c(-1, 0.1)),
  data = D1, family = nb(link = "log"), method = "REML"
)
fit2 <- gam(
  response ~ offset(offset0) +
    s(x, y, bs = "spdePC", xt = list(
      mesh = MISO_E13$meshes$spdePC_g999,
      pc_cutoff = 0.999
    ), sp = c(-1, 0.1)),
  data = D2, family = nb(link = "log"), method = "REML"
)

fit1$smooth[[1L]]$pc_retained_dimension
fit1$smooth[[1L]]$pc_score_Q
score_test(fit1, fit2, method = "liu")
```

Use the same pattern with `Visium_B`, `mt_co3`, and `mt_co2`. The two fitted
models passed to `score_test()` must use the same rows, mesh, and fixed
`kappa`.

## Run the packaged three-gene examples

The two scripts below each fit the bundled three genes with the baseline
`spde` smooth, use three SOCK workers, and evaluate the three explicit gene
pairs with `mgcvST.estimate()` followed by `mgcvST.test()`. They write the
fit diagnostics, pair results, and compact R objects to
`MGCVST_EXAMPLE_OUTPUT`.

```{r packaged-data-examples, eval=FALSE}
dir.create("mgcvST-results", showWarnings = FALSE)
Sys.setenv(
  MGCVST_EXAMPLE_OUTPUT = normalizePath("mgcvST-results")
)

source(system.file("examples", "MISO_E13.R", package = "mgcvST"))
source(system.file("examples", "Visium_B.R", package = "mgcvST"))
```

The high-throughput interface requires a positive-definite SPDE precision and
therefore currently supports the baseline `spde` smooth. Use the direct
`spdePC` workflow above for PC-truncated pair tests.

## Reproduce the bundled simulations

`inst/examples/null.R` performs the null tail/type-I simulation. Its defaults
are 100,000 simulations and 20 workers. `inst/examples/alternative.R` uses
100 replicates at each of 100 correlation values. Both compare the baseline
and `0.999` PC-truncated representations on the included MISO E13 and Visium
B examples and save their results under `MGCVST_EXAMPLE_OUTPUT`.

For a quick local smoke run, reduce the counts before sourcing either script.

```{r simulations, eval=FALSE}
Sys.setenv(
  MGCVST_EXAMPLE_OUTPUT = normalizePath("mgcvST-results"),
  MGCVST_EXAMPLE_WORKERS = 3,
  MGCVST_EXAMPLE_NULL_REPS = 300,
  MGCVST_EXAMPLE_POWER_REPS = 10
)

source(system.file("examples", "null.R", package = "mgcvST"))
source(system.file("examples", "alternative.R", package = "mgcvST"))
```

## Construct a mesh from a boundary

`spde_mesh()` accepts either a manually curated polygon boundary or a boundary
returned by `spde_boundary_learn()`. The example data use curated boundaries.

```{r mesh, eval=FALSE}
loc <- as.matrix(MISO_E13$covariates[, c("x", "y")])
boundary <- spde_boundary_learn(loc, n_holes = 0L)
mesh <- spde_mesh(boundary, loc = loc, target = 150L)
plot(mesh)
```
